---
title: "Siegle-Hw3"
author: "Aaron_Siegle"
date: "2026-09-27"
output: html_document
---

```{r, message=FALSE}
# load libraries
library(spdep)
library(sf)
library(terra)
library(spatialreg)
library(viridis)
library(data.table)
library(spectralGP)
library(RColorBrewer)
library(spgwr)
```


```{r}
# Import the data
pres_data_path <- "C:/Users/aasie/OneDrive/Documentos/GIS713-Homework3/Data/2016_US_County_Level_Presidential_Results.csv"
acs_county_data_path <- "C:/Users/aasie/OneDrive/Documentos/GIS713-Homework3/Data/acs2015_county_data.csv"
us_county_shp_path <- "C:/Users/aasie/OneDrive/Documentos/GIS713-Homework3/cb_2015_us_county_20m/cb_2015_us_county_20m.shp"
us_county <- st_read(us_county_shp_path)
election_results <- fread(pres_data_path)
acs_counties <- fread(acs_county_data_path)
head(election_results)
head(acs_counties)
head(us_county)
```
```{r}
# Look at column names in the datasets 
colnames(acs_counties)
colnames(election_results)
print(nrow(acs_counties))
print(nrow(election_results))
```

```{r}
# Remove non-CONUS counties from both datasets 
acs_counties <- acs_counties[
  !acs_counties$State %in% c("Alaska", "Hawaii", "Puerto Rico"), ]
election_results <- election_results[!state_abbr == "AK" & 
                                    !state_abbr == "HI" & 
                                    !state_abbr == "PR"]
print(nrow(acs_counties))
print(nrow(election_results))
```


```{r}
# Spatial join between the two datasets
election_counties <- merge(acs_counties,
                           election_results,
                           by.x="CensusId", 
                           by.y="combined_fips", 
                           all.x=TRUE, 
                           all.y=TRUE
                           )
head(election_counties)
colnames(election_counties)
```

```{r}
# Create new variables 
election_counties[, frac_non_citizen := 1 - (Citizen / TotalPop)][,
                  frac_women := Women/TotalPop][,
                  frac_nonwhite := 1 - (White/TotalPop)][,
                  frac_voting := total_votes/TotalPop]
head(election_counties)
```

```{r}
# Subset only to specified variables
election_counties_subset <- election_counties[, .SD, .SDcols = c(
    "CensusId", "State", "County", 
    "per_dem", "per_gop", "frac_voting", 
    "frac_women", "frac_nonwhite", "frac_non_citizen", 
    "Income", "Poverty", "ChildPoverty", "Professional"
    )]
colnames(election_counties_subset)
```

```{r}
# Subset only to counties where all variables are complete 
election_counties_complete <- election_counties_subset[
  complete.cases(election_counties_subset)
]
print(nrow(election_counties_subset))
print(nrow(election_counties_complete))
```

```{r}
# Question 2: Create a map with an appropriate divergent color ramp that shows the GOP vote fraction for each county
# Join in the spatial attributes from the shapefile
us_county$FIPS <- as.numeric(paste0(us_county$STATEFP, us_county$COUNTYFP))
us_county$FIPS <- as.integer(us_county$FIPS)
election_counties_complete <- as.data.frame(election_counties_complete)
acs_counties_spatial <- merge(
  us_county, 
  election_counties_complete,
  by.x = "FIPS", 
  by.y = "CensusId",
  all.x = TRUE,
  all.y = TRUE
)
head(acs_counties_spatial)
```

```{r}
plot(acs_counties_spatial["per_gop"], 
     breaks = seq(0,1, length.out=11),
     pal = rev(brewer.pal(10, "RdBu")), 
     main = "GOP Vote Fraction by County", 
     xlim = c(-125, -66), ylim = c(24, 50),
     border = NA
     )
```

```{r,warning=FALSE}
# Question 3: Explain the variation in county-level GOP vote share using the covariates. Test for spatial dependence and fit a spatial regression variant if necessary (use lm.morantest rather than moran.test).
# Build nearest neighbors 
nb <- poly2nb(acs_counties_spatial, queen = TRUE)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
summary(nb)
```

```{r}
# Using lm.morantest to test for spatial autocorrelation 
m0 <- lm(per_gop ~ 1, data = acs_counties_spatial)
lm.morantest(m0, lw, zero.policy = TRUE)

m1 <- lm(per_gop ~ frac_women + frac_nonwhite + frac_non_citizen + 
           frac_voting + Income + Poverty + ChildPoverty + Professional, 
         data = acs_counties_spatial)
ols_moran <- lm.morantest(m1, lw, zero.policy = TRUE)

ols_coefs <- summary(m1)$coefficients
ols_est <- ols_coefs[, "Estimate"]
ols_se <- ols_coefs[, "Std. Error"]
```

```{r}
# Moran's I is significant, which violates the assumption of spatial autocorrelation, meaning that we to choose a spatial regression variant 
lm.RStests(m1, lw, test= c("RSerr", "RSlag", "adjRSerr", "adjRSlag"), 
           zero.policy = TRUE)


# Running the spatial error model 
sem <- errorsarlm(formula(m1), data = acs_counties_spatial, listw = lw, 
                  zero.policy = TRUE)
summary(sem)
sem_coefs <- summary(sem)$Coef
sem_est <- sem_coefs[, "Estimate"]
sem_se <- sem_coefs[, "Std. Error"]

# Running the spatial lag model
slm <- lagsarlm(formula(m1), data = acs_counties_spatial, listw = lw, 
                  zero.policy = TRUE)
summary(slm)
slm_coefs <- summary(slm)$Coef
slm_est <- slm_coefs[, "Estimate"]
slm_se <- slm_coefs[, "Std. Error"]
```
Here we would use the spatial lag model because even though the two models have the same value p-value, the spatial lag model has the lower AIC value, which means that the spatial lag model has a higher accuracy when compared to the spatial error model. Both models have a high AIC because I used all of the other variables in the model as covariates, which made the model more complex. While this may improve the overall predictive power of the model, it reduces the incremental gain in model accuracy for each variable, thus inflating the AIC. 

```{r}
# Question 4: Create maps of the residuals for the OLS model as well as the two spatial variants. Create a table that compares the coefficient values and standard errors between the models.
# Data that OLS acts upon 
vars <- c("per_gop", "frac_women", "frac_nonwhite", "frac_non_citizen",
          "frac_voting", "Income", "Poverty", "ChildPoverty", "Professional")
dat <- acs_counties_spatial[complete.cases(st_drop_geometry(acs_counties_spatial)[, vars]), ]
dat <- st_make_valid(dat)

# Extract the residuals for each model 
dat$ols_resid <- as.numeric(resid(m1))
dat$sem_resid <- as.numeric(resid(sem))
dat$slm_resid <- as.numeric(resid(slm))

# Create three maps that are side by side 
lim <- max(abs(c(dat$ols_resid, 
                 dat$sem_resid,
                 dat$slm_resid)))
brks <- seq(-lim, lim, length.out = 11)
cols <- rev(brewer.pal(10, "RdBu"))

map_resid <- function(x, title) {
  plot(st_geometry(dat), 
       col = cols[cut(x, brks, include.lowest = TRUE)], 
       main = title,
       border = NA)
}

par(mfrow = c(3,1), mar= c(0,0,3,0))
map_resid(dat$ols_resid, "OLS Residuals")
map_resid(dat$sem_resid, "Spatial Error Model Residuals")
map_resid(dat$slm_resid, "Spatial Lag Model Residuals")

legend("left", 
       legend = levels(cut(0, brks, include.lowest = TRUE)), 
       fill = cols, cex = 0.7, bty = "n", title = "Residuals")

par(mfrow = c(1,1))
```

```{r}
# Create a table that compares the coefficient values and standard errors between the models.
coef_table <- data.frame(
  Term     = rownames(ols_coefs),
  OLS_Est  = ols_est, OLS_SE = ols_se,
  SEM_Est  = sem_est, SEM_SE = sem_se,
  SLM_Est  = slm_est, SLM_SE = slm_se,
  row.names = NULL
)
coef_table[, -1] <- round(coef_table[, -1], 4)
coef_table
```

```{r, warning=FALSE}
# Question 5: Fit a geographically weighted regression. Create maps w/ divergent color ramps that depict each of the effects on county-level GOP vote share.

pred_vars <- setdiff(vars, "per_gop")
f <- per_gop ~ frac_women + frac_nonwhite + frac_non_citizen +
  frac_voting + Income + Poverty + ChildPoverty + Professional

dat_proj <- st_transform(dat, 5070)
coords   <- st_coordinates(st_centroid(st_geometry(dat_proj)))

dat_df <- st_drop_geometry(dat)
dat_df[pred_vars] <- lapply(dat_df[pred_vars], function(x) as.numeric(scale(x)))

gop_gwr <- gwr(f, data = dat_df, coords = coords,
               adapt = 0.05,
               hatmatrix = FALSE, longlat = FALSE)
saveRDS(gop_gwr, "gop_gwr.rds")

```


```{r}
# Create maps w/ divergent color ramps that depict each of the effects on county-level GOP vote share.
res <- as.data.frame(gop_gwr$SDF)
sapply(res[, pred_vars], function(x) sum(is.finite(x)))   

for (v in pred_vars) {
  b <- res[[v]]
  b[!is.finite(b)] <- NA
  dat[[paste0("b_", v)]] <- b
  lim <- max(abs(b), na.rm = TRUE)

  plot(dat[paste0("b_", v)], border = NA,
       breaks = seq(-lim, lim, length.out = 11),
       pal = rev(brewer.pal(10, "RdBu")),
       main = paste("Effect of", v, "on per_gop"))}
  
```

